---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# caribouMetrics

<!-- badges: start -->
[![R-CMD-check](https://github.com/LandSciTech/caribouMetrics/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/LandSciTech/caribouMetrics/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

caribouMetrics provides implementations of several different models. Firstly, it
implements the caribou resource selection probability functions described in
Hornseth and Rempel (2016) *Seasonal resource selection of woodland caribou (Rangifer tarandus caribou) across a gradient of anthropogenic disturbance*. This allows for a spatial prediction of caribou habitat use across 13 caribou ranges in Ontario. The package also includes an implementation of the population and demographic models in *Science to inform policy: linking population dynamics to habitat for a threatened species in Canada* by Johnson et. al. (2020). These functions allow users to calculate metrics of disturbance, predict demographic rates for a given level of disturbance and simulate population growth over time.  

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("LandSciTech/caribouMetrics")
```
## Example

This is a basic example which shows you how to use the main functions:

```{r example}
library(caribouMetrics)

pthBase <- system.file("extdata", package = "caribouMetrics")

# load example data
landCoverD <- raster::raster(file.path(pthBase, "landCover.tif")) 
  # convert PLC classes to resource types used in the model 
landCoverD <- reclassPLC(landCoverD)
eskerDras <- raster::raster(file.path(pthBase, "eskerTif.tif"))
eskerDshp <- sf::read_sf(file.path(pthBase, "esker.shp"))
natDistD <- raster::raster(file.path(pthBase, "natDist.tif"))
anthroDistD <-raster::raster(file.path(pthBase, "anthroDist.tif"))
linFeatDras <- raster::raster(file.path(pthBase, "linFeatTif.tif"))
projectPolyD <- sf::read_sf(file.path(pthBase, "projectPoly.shp"))

# Calculate habitat use
carHab1 <- caribouHabitat(
  landCover = landCoverD,
  esker = eskerDras, 
  natDist = natDistD, 
  anthroDist = anthroDistD, 
  linFeat = linFeatDras, 
  projectPoly = projectPolyD,
  caribouRange = "Churchill"
)

# plot the results
plot(carHab1)

# calculate disturbance 
disturb <- disturbanceMetrics(landCover = landCoverD,
                              linFeat = linFeatDras,  
                              natDist = natDistD,
                              projectPoly = projectPolyD)

results(disturb)

# Calculate demographic rates
demCoefs <- demographicCoefficients(replicates = 10)

demRates <- demographicRates(covTable = results(disturb),
                             popGrowthPars = demCoefs)
demRates

# Simulate population growth
popGrow <- popGrowthJohnson(N = 2000, numSteps = 20, R_bar = demRates$R_bar, 
                            S_bar = demRates$S_bar)

popGrow

```

More detailed examples are provided in the vignettes.

# References

Johnson, C.A., Sutherland, G.D., Neave, E., Leblond, M., Kirby, P., Superbie, C.
and McLoughlin, P.D., 2020. Science to inform policy: linking population
dynamics to habitat for a threatened species in Canada. Journal of Applied
Ecology, 57(7), pp.1314-1327.
<https://doi.org/10.1111/1365-2664.13637>

Hornseth, M.L. and Rempel, R.S., 2016. Seasonal resource selection of
woodland caribou (Rangifer tarandus caribou) across a gradient of
anthropogenic disturbance. Canadian Journal of Zoology, 94(2), pp.79-93.
https://doi.org/10.1139/cjz-2015-0101

## License
caribouMetrics - Terms and Condition of Use

Unless otherwise noted, source code of the `caribouMetrics` R package is covered under
Crown Copyright, Government of Canada, and distributed under the GPL3 license.

Copyright (C) Her Majesty the Queen in Right of Canada as represented by the Minister of the Environment 2021/(C) Sa Majesté la Reine du chef du Canada représentée par le ministre de l'Environnement 2021.
