---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# caribouMetrics

<!-- badges: start -->
<!-- badges: end -->

caribouMetrics is built off of two different models, one calculates predictors
described in Table 52 of Environment Canada (2011) Scientific Assessment to
Inform the Identification of Critical Habitat for Woodland Caribou and the other
implements the caribou resource selection probability functions described in
Rempel, R. and M. Hornseth. 2018. Range-specific seasonal resource selection
probability functions for 13 caribou ranges in Northern Ontario. The second part
of the package is the actively developed part and the rest of the README focuses
on it.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("LandSciTech/caribouMetrics")
```
## Example

This is a basic example which shows you how to use the primary function:

```{r example}
#library(caribouMetrics)
devtools::load_all()
pthBase <- "tests/testthat/data/"

# load example data
plcD = raster(paste0(pthBase, "plc", ".tif"))
eskerDras = raster(paste0(pthBase, "eskerTif", ".tif"))
eskerDshp = st_read(paste0(pthBase, "esker", ".shp"), quiet = TRUE)
friD = raster(paste0(pthBase, "fri", ".tif"))
ageD = raster(paste0(pthBase, "age", ".tif"))
natDistD = raster(paste0(pthBase, "natDist", ".tif"))
anthroDistD = raster(paste0(pthBase, "anthroDist", ".tif"))
harvD = raster(paste0(pthBase, "harv", ".tif"))
linFeatDras = raster(paste0(pthBase, "linFeatTif", ".tif"))
projectPolyD = st_read(paste0(pthBase, "projectPoly", ".shp"), quiet = TRUE)

# Calculate habitat use
carHab1 <- caribouHabitat(
  plc = plcD, esker = eskerDras, fri = friD, age = ageD, natDist = natDistD, 
  anthroDist = anthroDistD, harv = harvD,
  linFeat = linFeatDras, projectPoly = projectPolyD,
  friLU = read.csv(paste0(pthBase, "friLU", ".csv"), stringsAsFactors = FALSE) %>%
    mutate(RFU = toupper(RFU) %>% stringr::str_replace("HRDMW", "HRDMX")),
  caribouRange = "Churchill", 
  winArea = 500
)

# plot the results
plot(carHab1)
```

More detailed examples are provided in the vignettes/overall_vignette.Rmd and in scripts/notesOnHandRModel.Rmd
