---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# caribouMetrics

<!-- badges: start -->
<!-- badges: end -->

caribouMetrics is built off of two different models, one calculates predictors
described in Table 52 of Environment Canada (2011) Scientific Assessment to
Inform the Identification of Critical Habitat for Woodland Caribou and the other
implements the caribou resource selection probability functions described in
Hornseth and Rempel (2016) Seasonal resource selection of woodland 
caribou (*Rangifer tarandus caribou*) across a gradient of anthropogenic disturbance. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("LandSciTech/caribouMetrics")
```
## Example

This is a basic example which shows you how to use the main functions:

```{r example}
library(caribouMetrics)

pthBase <- system.file("extdata", package = "caribouMetrics")

# load example data
landCoverD <- raster::raster(file.path(pthBase, "landCover.tif")) 
  # convert PLC classes to resource types used in the model 
landCoverD <- reclassPLC(landCoverD)
eskerDras <- raster::raster(file.path(pthBase, "eskerTif.tif"))
eskerDshp <- sf::read_sf(file.path(pthBase, "esker.shp"))
natDistD <- raster::raster(file.path(pthBase, "natDist.tif"))
anthroDistD <-raster::raster(file.path(pthBase, "anthroDist.tif"))
linFeatDras <- raster::raster(file.path(pthBase, "linFeatTif.tif"))
projectPolyD <- sf::read_sf(file.path(pthBase, "projectPoly.shp"))

# Calculate habitat use
carHab1 <- caribouHabitat(
  landCover = landCoverD,
  esker = eskerDras, 
  natDist = natDistD, 
  anthroDist = anthroDistD, 
  linFeat = linFeatDras, 
  projectPoly = projectPolyD,
  caribouRange = "Churchill"
)

# plot the results
plot(carHab1)

# calculate disturbance 
disturb <- disturbanceMetrics(landCover = landCoverD,
                              linFeat = linFeatDras,  
                              natDist = natDistD,
                              projectPoly = projectPolyD)

results(disturb)

# Calculate demographic rates
demCoefs <- demographicCoefficients(replicates = 10)

demRates <- demographicRates(covTable = results(disturb),
                             popGrowthPars = demCoefs)
demRates

# Simulate population growth
popGrow <- popGrowthJohnson(N = 2000, numSteps = 20, Rec_bar = demRates$R_bar, 
                            S_bar = demRates$S_bar)

popGrow

```

More detailed examples are provided in the vignettes.