---
title: "workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
#| width: 100
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = FALSE
)
```

<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{mermaid}
%%| column: screen
%%| fig.width: 20
---
displayMode: compact
---
%%{init: { "gantt": { "barGap": 20 } } }%%
gantt
    title Paper workflow
    dateFormat YYYY-MM-DD
    axisFormat %i
    
    section A
    getSimsInitial :a1, 2000-01-01, 20d
    .              :done, ax, after a1, 1d  
    runScnSet      :a2, after ax, 245d
    .              :done, ay, after a2, 1d  
    plotRes        :a3, after ay, 20d  
    
    section B
    getScenarioDefaults  :b2, after ax, 40d
    .                    :done, bx, after b2, 1d  
    simulateObservations :b3, after bx, 80d
    .                    :done, by, after b3, 1d 
    caribouBayesianPM    :b4, after by, 101d
    .                    :done, bz, after b4, 1d  
    getOutputTables      :b5, after bz, 20d
    
    section C
    trajectoriesFrom??  :c1, after bx, 40d

    bbouMakeSummaryTable     :c5, after by, 40d
    .                        :done, cy, after c5, 1d
    getSimsInitial           :c6, after cy, 40d
    
    section D
    getPriors    :d4, after by, 20d
    
```


```{mermaid}
%%| column: screen
%%| fig.width: 20
---
displayMode: compact
---
%%{init: { "gantt": { "barGap": 20 } } }%%
gantt
    title Paper workflow
    dateFormat YYYY-MM-DD
    axisFormat %i
    
    section A
    getSimsInitial :a1, 2000-01-01, 20d
    .              :done, ax, after a1, 1d  
    runScnSet      :a2, after ax, 245d
    .              :done, ay, after a2, 1d  
    plotRes        :a3, after ay, 20d  
    
    section B
    getSimsNational      :b1, 2000-01-01, 20d
    
    getScenarioDefaults  :b2, after ax, 20d
    .                    :done, bx, after b2, 1d  
    simulateObservations :b3, after bx, 101d
    .                    :done, by, after b3, 1d 
    caribouBayesianPM    :b4, after by, 101d
    .                    :done, bz, after b4, 1d  
    getOutputTables      :b5, after bz, 20d
    
    section C
    demographicCoefficients  :c1, after bx, 40d
    .                        :done, cx, after c1, 1d  
    demographicRates         :c2, after cx, 20d
    compositionBiasCorr      :c3, after c2, 20d 
    caribouPopGrowth         :c4, after c3, 20d
    
    bbouMakeSummaryTable     :c5, after by, 20d
    .                        :done, cy, after c5, 1d
    getSimsInitial           :c6, after cy, 80d
    
    section D
    getCoefs     :d1, after bx, 20d
    sampleCoefs  :d2, after d1, 20d
    
    sampleRates  :d3, after cx, 20d 
    
    getPriors    :d4, after by, 20d
    
    compositionBiasCorr       :d5, after cy, 20d
    caribouPopSimMCMC         :d6, after d5, 20d
    convertTrajectories       :d7, after d6, 20d
    summarizeTrajectories     :d8, after d7, 20d
    
```


```{r eval=FALSE}
# don't run just used to create chart with new names

lu_tbl <- clipr::read_clip_tbl()

gantt_str <- clipr::read_clip()

new_gantt <- stringr::str_replace_all(gantt_str, setNames(lu_tbl$NewFunctionName, lu_tbl$FunctionName))

cat(new_gantt, sep = "\n")
```

```{mermaid}
%%| column: screen
%%| fig.width: 20
---
displayMode: compact
---
%%{init: { "gantt": { "barGap": 20 } } }%%

gantt
    title Paper workflow new names
    dateFormat YYYY-MM-DD
    axisFormat %i
    
    section A
    trajectoriesFromAny :a1, 2000-01-01, 25d
    .              :done, ax, after a1, 1d  
    bayesianScenariosWorkflow      :a2, after ax, 305d
    .              :done, ay, after a2, 1d  
    plotCompareTrajectories        :a3, after ay, 25d  
    
    section B
    trajectoriesFromNtnl*      :b1, 2000-01-01, 25d
    
    getScenarioDefaults  :b2, after ax, 25d
    .                    :done, bx, after b2, 1d  
    simulateObservations :b3, after bx, 126d
    .                    :done, by, after b3, 1d 
    bayesianTrajectoryWorkflow    :b4, after by, 126d
    .                    :done, bz, after b4, 1d  
    compareTrajectories      :b5, after bz, 25d
    
    section C
    getNationalCoefficients :c1, after bx, 50d
    .                       :done, cx, after c1, 1d  
    estimateNationalRates   :c2, after cx, 25d
    compositionBiasCorr*    :c3, after c2, 25d 
    caribouPopGrowth        :c4, after c3, 25d
    
    estimateBayesianRates :c5, after by, 25d
    .                     :done, cy, after c5, 1d
    trajectoriesFromAny   :c6, after cy, 100d
    
    section D
    subsetNationalCoefs   :d1, after bx, 25d
    sampleNationalCoefs   :d2, after d1, 25d
    
    estimateNationalRate  :d3, after cx, 25d 
    
    betaNationalPriors    :d4, after by, 25d
    
    compositionBiasCorr*      :d5, after cy, 25d
    simTrajFromPosterior*     :d6, after d5, 25d
    convertTrajectories       :d7, after d6, 25d
    summarizeTrajectories     :d8, after d7, 25d
```


# Getting demographic rates, trajectories, and trajectory summaries

```{mermaid}
%%| column: screen
%%| fig.width: 20
---
displayMode: compact
---
%%{init: { "gantt": { "barGap": 20 } } }%%
gantt
    title Demographic rates, trajectories, and trajectory summaries from the national model. 
    dateFormat YYYY-MM-DD
    axisFormat %i
    
    section A
    getSimsInitial :a1, 2000-01-01, 240d

    section B
    getSimsNational       :b1, 2000-01-01, 160d
    convertTrajectories     :b2, after b1, 40d
    summarizeCaribouPopSim     :b3, after b2, 40d

    section C
    demographicCoefficients  :c1, 2000-01-01, 40d
    demographicRates         :c2, after c1, 40d
    compositionBiasCorr      :c3, after c2, 40d 
    caribouPopGrowth         :c4, after c3, 40d

```

Note that if a fitted disturbance scenario is not specified (either as a disturbance table or disturbance scenario parameters) getSimsInitial returns demographic summaries for different levels of anthropogenic disturbance from 1 to 100. Otherwise, outputs include demographic trajectories and summaries of those trajectories.


```{mermaid}
%%| column: screen
%%| fig.width: 20
---
displayMode: compact
---
%%{init: { "gantt": { "barGap": 20 } } }%%
gantt
    title Demographic rates, trajectories, and trajectory summaries from Bayesian model posterior samples. 
    dateFormat YYYY-MM-DD
    axisFormat %i
    
    section A
    bbouMakeSummaryTable :a1, 2000-01-01, 40d
    .              :done, ax, after a1, 1d  
    getSimsInitial :a2, after ax, 160d

    section B
    compositionBiasCorr :b1, after ax, 40d
    caribouPopSimMCMC  :b2, after b1, 40d
    convertTrajectories :b3, after b2, 40d
    summarizeCaribouPopSim  :b4, after b3, 40d

```

TO DO: Demographic rates, trajectories, and trajectory summaries from Bayesian model posterior summaries...

\* indicates abbreviated function names for trajectoriesFromNational, compositionBiasCorrection and simulateTrajectoriesFromPosterior.

Using grey "done" period to create some space between groupings of functions 
