---
title: "Bayesian Demographic Projection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Demographic Projection}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 3.5
)
```

```{r setup}
library(caribouMetrics)
library(ggplot2)
library(dplyr)

theme_set(theme_bw())
```

caribouMetrics provides a simple Bayesian integrated population model that integrates prior information from Johnson et al.'s (2020) national analysis of demographic-disturbance relationships with available local demographic data to project population growth. In addition, methods are provided for simulating local population dynamics and monitoring programs. 

Boreal caribou monitoring programs typically involve marking caribou with  telemetry GPS/VHF collars which are used to monitor caribou over time and detect the death of marked animals. This data is then used to estimate survival, while recruitment is estimated by locating collared individuals using aircraft and then surveying the entire group of caribou to estimate the ratio of cows:calves in the group. Because this estimate is based on more than just the sample of collared cows, the sample size for recruitment estimates is typically larger than for survival. 

## Integration of local demographic data and national disturbance-demographic relationships in a Bayesian population model

Following Eacker et al. (2019), the observed number
of calves $\hat{J}_t$ is a binomially distributed function of the
observed number of adult female caribou $\hat{W}_t$ and the estimated
recruitment rate $R_t$:
$$\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).$$

We model recruitment probability as a function of anthropogenic disturbance and
fire with a log link (Dyson et al., 2022; Johnson et al., 2020; Stewart et al.,
2023) and Gaussian distributed random year effect (Eacker et al., 2019):
$$\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t
\sim \text{Normal}(0,\sigma^2_{R}).$$
Estimated recruitment probability is adjusted for sex ratio,
misidentification biases, and age of first reproduction
(DeCesare et al., 2012; Eacker et al., 2019) to get expected
recruits per female:
$$X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.$$ 
The composition survey bias correction term $c$ depends on the apparent number
of adult females per collared animal $w$, the ratio of young bulls to adult
females $q$, the probability of misidentifying young bulls as adult females and
vice versa $u$, and the probability of missing a calf $z$ (see
compositionBiasCorrection()). To maintain consistency with Johnson et al.
(2020) we also provide an option (`adjustR = FALSE`) to only
adjust for sex ratio and misidentification biases $X_t=cR_t/2$.

In the Bayesian model, prior uncertainty about the value of the bias
correction term $c$ is approximated with a Log-normal distribution.
Given the apparent number of adult females per collared
animal $w$, the mean and standard deviation of 10,000 samples of $\log{c}$
is calculated with the default being to assume the ratio of young
bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
misidentification probability $u$ varies uniformly between 0 and 0.2,
and the probability of missing a calf $z$ varies uniformly between 0 and
0.2 (see getPriors).

Also following Eacker et al. (2019), Kaplan-Meier estimates of observed adult
female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are
estimated from known-fate radio collar data using the
[survival](https://CRAN.R-project.org/package=survival) package, with precisions
for years with no observed mortality estimated using a separate Bayesian model.
Observed estimated survival is a Gaussian distributed function of the true
survival probability: 
$$\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).$$ 
As in the national demographic model, we model survival probability as an
adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
(Dyson et al., 2022; Johnson et al., 2020; Stewart et al., 2023) and Gaussian
distributed random year effect (Eacker et al., 2019):
$$S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).$$
To handle cases in which there is little or no survival data, we also
implemented an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
(Schaub and Kery, 2021):
$$\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).$$ 
The exponential survival method can be used in all cases instead of the
Kaplan-Meier method by setting `survAnalysisMethod = "Exponential"`.

All regression coefficients are assumed to be Gaussian distributed, with priors
derived from the expected values and 95% confidence intervals estimated by
Johnson et al (2020) (`popGrowthTableJohnsonECCC`). We calibrated the prior
distributions so that the 95% prior prediction intervals for survival and
recruitment from the Bayesian model match the range between the 2.5% and 97.5%
quantiles of 1000 simulated survival and recruitment trajectories (see
`getPriors()`).

### Using real observed data
caribouMetrics includes example csv files of collar survival and calf cow count data as well as disturbance data that can be used as templates for the format for observed data. 
```{r load-obs}
survObs <- read.csv(system.file("extdata/simSurvData.csv", package = "caribouMetrics"))
ageRatioObs <- read.csv(system.file("extdata/simAgeRatio.csv", package = "caribouMetrics"))
distObs <- read.csv(system.file("extdata/simDisturbance.csv", package = "caribouMetrics"))

survObs %>% group_by(Year) %>% 
  summarise(n_collars = n(),
            n_die = sum(event)) %>% 
  ggplot(aes(Year))+
  geom_line(aes(y = n_collars))+
  geom_col(aes(y = n_die))+
  labs(y = "Number of Collars (line) and deaths (bars)")

ageRatioObs %>% 
  ggplot(aes(Year, Count, fill = Class))+
  geom_col(position = "dodge")

distObs %>% ggplot(aes(Year, Anthro))+
  geom_line()

```
Plotting the data shows that the collaring program included 60 individuals with collars being replenished every 4 years over a monitoring period of 15 years. Calf cow surveys typically counted ~150 cows and ~50 calves. There is no disturbance in the years of the observation data but disturbance is expected to increase quickly in the future.

These data sets can be supplied to `caribouBayesianIPM()`to project the impact on the caribou population as disturbance increases over the next 20 years. 

```{r mod-real}
mod_real <- caribouBayesianIPM(survData = survObs, ageRatio = ageRatioObs, 
                              disturbance = distObs, 
                              # only set to speed up vignette. Normally keep defaults.
                              Niter = 150, Nburn = 100)

str(mod_real, max.level = 2)
```

The returned object contains an `rjags` object and a list with the modified input data. We can get tables summarizing the results using `getOutputTables()`.
```{r}
mod_tbl <- getOutputTables(mod_real)
str(mod_tbl)
```

And plot the results with `plotRes()`.
```{r}
plotRes(mod_tbl, "Recruitment", labFontSize = 10)
```

We can also compare our local observed data to what would be projected by the national model with out considering local population specific data. 
```{r}
simNational <- getSimsNational()

mod_nat_tbl <- getOutputTables(mod_real, 
                               simNational = simNational,
                               getKSDists = FALSE)

plotRes(mod_nat_tbl,
        c("Recruitment", "Adult female survival", "Population growth rate"),
        labFontSize = 10)

```

From these graphs we can see that this local population seems to have slightly higher demographic rates than would have been predicted by the national model alone and that the uncertainty around the predictions is lower when the local observations are included. Note that the population's response to anthropogenic disturbance in completely determined by the national model since there was 0% disturbance during the observation period.  

## Simulation of local population dynamics and monitoring

To simulate plausible demographic trajectories for example populations
we used a modified version of Johnson et al's (2020)
two-stage demographic model described by Dyson et al.
(2022). The true number of post-juvenile females that
survive from year $t$ to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum
potential recruitment rate is adjusted for sex ratio, misidentification
biases and (optionally) delayed age at first reproduction
(DeCesare et al., 2012; Eacker et al., 2019):
$$\dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}, $$
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $q$ is the ratio of young bulls to adult females, $u$
is the probability of misidentifying young bulls as adult females and
vice versa, and $z$ is the probability of missing a calf.

Realized recruitment rate varies with population density
(Lacy et al., 2017), and the number of juveniles recruiting to the
post-juvenile class at the census is a binomially distributed function
of the number of surviving post-juvenile females and the adjusted
recruitment rate:
$$\dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).$$
Given default parameters, recruitment rate is lowest
$(0.5\dot{X}_t)$ when $\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$
at intermediate population sizes, and declines to $0.6\dot{X}_t$ as the
population reaches carrying capacity of $k=100$ times the initial
population size. The post-juvenile female population in the next year
includes both survivors and new recruits:
$\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t)$.

Interannual variation in survival and recruitment is modelled using
truncated beta distributions (rtrunc function;
Novomestky and Nadarajah (2016)):
$\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,\nu_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,\nu_S,l_S,h_S)$.
Coefficients of variation among years $(\nu_R,\nu_S)$ and
maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival
have default values set in `caribouPopGrowth()`. Expected recruitment ($\bar{R}_t$) and survival
($\bar{S}_t$) vary with disturbance according the beta regression models
estimated by Johnson et al. (2020:
$$ \bar{R}_t \sim Beta(\mu^R_t,\phi^R);log(\mu^R_t)=\dot{\beta^R_0}+\dot{\beta^R_a}A_t+\dot{\beta}^R_fF_t,$$
$$\bar{S}_t \sim (46\times Beta(\mu^S_t,\phi^S)-0.5)/45;log(\mu^S_t)=\dot{\beta^S_0}+\dot{\beta^S_a}A_t.$$
$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors (Ferrari and Cribari-Neto, 2004). At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see `demographicRates()`) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example among populations.

### Observed survival of collared animals

In our simulated populations, mortality risk does not vary over time or
among individuals, so the observed state (dead or alive) of a collared
animal $i$ in month $m$ and year $t$ depends on the true survival rate
$\dot{S}_t$ as follows:
$$\hat{I}_{i,t,m} \sim \text{Bernoulli}(\dot{S}^{1/12}_t\hat{I}_{i,t,m-1}).$$

### Observed recruitment from composition surveys

The number of cows in the composition survey $\hat{W}_t$ is given by the
number of collared cows $\hat{T}_t$ and the apparent number of adult
females per collared female observed in the composition survey $w$:
$$\hat{W}_t = w\hat{T}_t.$$ The number of observed calves $\hat{J}_t$
also depends on the unadjusted apparent recruitment for the population
$\dot{R}_t$: $$\hat{J}_t \sim \text{Binomial}(\hat{W}_t,\dot{R}_t).$$


To run the simulations we need to supply parameters that determine the disturbance scenario, the trajectory of the true population relative to the national model mean, and the collaring program details. All these parameters are set with `getScenarioDefaults()` which will create a table with the default values of all parameters and override the defaults for any values that are supplied. Below we define a scenario where we have 20 years of observations and 20 years of projection, increasing anthropogenic disturbance over time, and 30 collars deployed every year. We assume that 2 cows will be observed in aerial surveys for every collared cow. The default values are set for our simulated true population meaning that we assume the population has the same response to disturbance as the national model and that the population demographic rates are close to the national average.
See `getScenarioDefaults()` for a detailed description of each parameter.

```{r sim-obs}
scn_params <- getScenarioDefaults(
  # Anthropogenic disturbance increases by 2% per year in observation period and
  # 3% per year in projection period
  obsAnthroSlope = 2, projAnthroSlope = 3,
  # 20 years each of observations and projections
  obsYears = 20, projYears = 20,
  # Collaring program aims to keep 30 collars active
  collarCount = 30,
  # Collars are topped up every year
  collarInterval = 1,
  # Assume will see 2 cows in aerial survey for every collar deployed
  cowMult = 2
  )

scn_params

sim_obs <- simulateObservations(
  scn_params, 
  # collars fall off after 4 years and are deployed in May and fall off in August
  collarNumYears = 4, collarOffTime = 8, collarOnTime = 5, 
  printPlot = TRUE)


sim_obs$simSurvObs %>% group_by(Year) %>% 
  summarise(ncollar = n(), ndeaths = sum(event), 
            ndropped = sum(exit == 5 & event == 0),
            nadded = sum(enter == 7), 
            survsCalving = sum(exit >= 6)) %>% 
  ggplot(aes(Year))+
  geom_line(aes(y = ncollar))+
  geom_col(aes(y = ndeaths))+
  labs(y = "Number of Collars (line) and deaths (bars)")

sim_obs$ageRatioOut %>% 
  ggplot(aes(Year, Count, fill = Class))+
  geom_col(position = "dodge")

sim_obs$simDisturbance %>% ggplot(aes(Year, Anthro))+
  geom_line()
```

We can provide the simulated observations to `caribouBayesianIPM()` to project the population growth over time. This time we supply the expected true population metrics as well as the model results to `getOutputTables()` so that we can see how well our monitoring program captured the true population. 

```{r mod-sim-obs}
mod_sim <- caribouBayesianIPM(survData = sim_obs$simSurvObs,
                              ageRatio = sim_obs$ageRatioOut, 
                              disturbance = sim_obs$simDisturbance, 
                              # only set to speed up vignette. Normally keep defaults.
                              Niter = 150, Nburn = 100)

mod_sim_tbl <- getOutputTables(mod_sim, exData = sim_obs$exData, 
                               paramTable = sim_obs$paramTable,
                               simNational = simNational,
                               getKSDists = FALSE)

plotRes(mod_sim_tbl,
        c("Recruitment", "Adult female survival", "Population growth rate"),
        labFontSize = 10)
```


### Comparing many scenarios

```{r many-scns, fig.height=8, fig.width=7}
eParsIn <- list()
eParsIn$collarOnTime <- 1
eParsIn$collarOffTime <- 12
eParsIn$collarNumYears <- 3

# adjust recruitment for delayed age of first reproduction
adjustR <- T 

simBig <- getSimsNational(adjustR = adjustR)

scns <- expand.grid(
  obsYears = 10, projYears = 10, collarCount = 100, cowMult = 2, collarInterval = 2,
  assessmentYrs = 1, iAnthro = 0, obsAnthroSlope = 0, projAnthroSlope = 0, 
  sQuantile = c(0.1, 0.5, 0.9), rQuantile = c(0.1, 0.5, 0.9), N0 = 1000
)
scResults <- runScnSet(
  scns, eParsIn, simBig, getKSDists = FALSE,
  # only set to speed up vignette. Normally keep defaults.
  Niter = 150, Nburn = 100)

plotRes(scResults,
        c("Recruitment", "Adult female survival", "Population growth rate"), 
        facetVars = c("rQuantile", "sQuantile"))

```

### Modifying Bayesian model priors
By default `caribouBayesianIPM()` calls `getPriors()` internally to set the priors for the Bayesian model based on the national model and default uncertainty modifiers that have been calibrated to fit the national model while allowing for deviations based on local data. The parameters to `getPriors()` affect the level of uncertainty around the coefficients from the national model. If uncertainty in the national model is higher then the projections can deviate more from the national model if the observed data does. See `getPriors()` for details.  

### Troubleshooting
The national model results are cached if the default values are used. This cache can be updated by running `getSimsNational(forceUpdate = TRUE)`




