---
title: "Bayesian Demographic Projection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Demographic Projection}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 3.5
)
```

```{r setup}
library(caribouMetrics)
library(ggplot2)
library(dplyr)

theme_set(theme_bw())
```

caribouMetrics provides a simple Bayesian population model that integrates prior information from Johnson et al.'s (2020) national analysis of demographic-disturbance relationships with available local demographic data to project population growth. In addition, methods are provided for simulating local population dynamics and monitoring programs. 

The model is described in Hughes et al. Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions.

### Using real observed data
caribouMetrics includes example csv files of collar survival and calf cow count data as well as disturbance data that can be used as templates for the format for observed data. 
```{r load-obs}
survObs <- read.csv(system.file("extdata/simSurvData.csv", package = "caribouMetrics"))
ageRatioObs <- read.csv(system.file("extdata/simAgeRatio.csv", package = "caribouMetrics"))
distObs <- read.csv(system.file("extdata/simDisturbance.csv", package = "caribouMetrics"))

survObs %>% group_by(Year) %>% 
  summarise(n_collars = n(),
            n_die = sum(event)) %>% 
  ggplot(aes(Year))+
  geom_line(aes(y = n_collars))+
  geom_col(aes(y = n_die))+
  labs(y = "Number of Collars (line) and deaths (bars)")

ageRatioObs %>% 
  ggplot(aes(Year, Count, fill = Class))+
  geom_col(position = "dodge")

distObs %>% ggplot(aes(Year, Anthro))+
  geom_line()

```
Plotting the data shows that the collaring program included 60 individuals with collars being replenished every 3 years over a monitoring period of 15 years. Calf cow surveys typically counted ~150 cows and ~50 calves. There is no disturbance in the years of the observation data but disturbance is expected to increase quickly in the future.

These data sets can be supplied to `caribouBayesianPM()`to project the impact on the caribou population as disturbance increases over the next 20 years. 

```{r mod-real}
mod_real <- caribouBayesianPM(survData = survObs, ageRatio = ageRatioObs, 
                              disturbance = distObs, 
                              # only set to speed up vignette. Normally keep defaults.
                              Niter = 150, Nburn = 100)

str(mod_real, max.level = 2)
```

The returned object contains an `rjags` object and a list with the modified input data. We can get tables summarizing the results using `getOutputTables()`.
```{r}
mod_tbl <- getOutputTables(mod_real)
str(mod_tbl)
```

And plot the results with `plotRes()`.
```{r}
plotRes(mod_tbl, "Recruitment", labFontSize = 10)
```

We can also compare our local observed data to what would be projected by the national model with out considering local population specific data. 
```{r}
simNational <- getSimsNational()

mod_nat_tbl <- getOutputTables(mod_real, 
                               simNational = simNational,
                               getKSDists = FALSE)

plotRes(mod_nat_tbl,
        c("Recruitment", "Adult female survival", "Population growth rate"),
        labFontSize = 10)

```

From these graphs we can see that this local population seems to have slightly higher demographic rates than would have been predicted by the national model alone and that the uncertainty around the predictions is lower when the local observations are included. Note that the population's response to anthropogenic disturbance is completely determined by the national model since there was 0% disturbance during the observation period.  

## Simulation of local population dynamics and monitoring

To run the simulations we need to supply parameters that determine the disturbance scenario, the trajectory of the true population relative to the national model mean, and the collaring program details. All these parameters are set with `getScenarioDefaults()` which will create a table with the default values of all parameters and override the defaults for any values that are supplied. Below we define a scenario where we have 20 years of observations and 20 years of projection, increasing anthropogenic disturbance over time, and 30 collars deployed every year. We assume that 2 cows will be observed in aerial surveys for every collared cow. The default values are set for our simulated true population meaning that we assume the population has the same response to disturbance as the national model and that the population demographic rates are close to the national average.
See `getScenarioDefaults()` for a detailed description of each parameter.

```{r sim-obs}
scn_params <- getScenarioDefaults(
  # Anthropogenic disturbance increases by 2% per year in observation period and
  # 3% per year in projection period
  obsAnthroSlope = 2, projAnthroSlope = 3,
  # 20 years each of observations and projections
  obsYears = 20, projYears = 20,
  # Collaring program aims to keep 30 collars active
  collarCount = 30,
  # Collars are topped up every year
  collarInterval = 1,
  # Assume will see 2 cows in aerial survey for every collar deployed
  cowMult = 2
  )

scn_params

sim_obs <- simulateObservations(
  scn_params, 
  # collars fall off after 4 years and are deployed in May and fall off in August
  collarNumYears = 4, collarOffTime = 8, collarOnTime = 5, 
  printPlot = TRUE)

sim_obs$simSurvObs %>% group_by(Year) %>% 
  summarise(ncollar = n(), ndeaths = sum(event), 
            ndropped = sum(exit == 5 & event == 0),
            nadded = sum(enter == 7), 
            survsCalving = sum(exit >= 6)) %>% 
  ggplot(aes(Year))+
  geom_line(aes(y = ncollar))+
  geom_col(aes(y = ndeaths))+
  labs(y = "Number of Collars (line) and deaths (bars)")

sim_obs$ageRatioOut %>% 
  ggplot(aes(Year, Count, fill = Class))+
  geom_col(position = "dodge")

sim_obs$simDisturbance %>% ggplot(aes(Year, Anthro))+
  geom_line()
```

We can provide the simulated observations to `caribouBayesianPM()` to project the population growth over time. This time we supply the expected true population metrics as well as the model results to `getOutputTables()` so that we can see how well our monitoring program captured the true population. 

```{r mod-sim-obs}
mod_sim <- caribouBayesianPM(survData = sim_obs$simSurvObs,
                              ageRatio = sim_obs$ageRatioOut, 
                              disturbance = sim_obs$simDisturbance, 
                              # only set to speed up vignette. Normally keep defaults.
                              Niter = 150, Nburn = 100)

mod_sim_tbl <- getOutputTables(mod_sim, exData = sim_obs$exData, 
                               paramTable = sim_obs$paramTable,
                               simNational = simNational,
                               getKSDists = FALSE)

plotRes(mod_sim_tbl,
        c("Recruitment", "Adult female survival", "Population growth rate"),
        labFontSize = 10)
```

### Comparing many scenarios

```{r many-scns, fig.height=8, fig.width=7}
eParsIn <- list()
eParsIn$collarOnTime <- 1
eParsIn$collarOffTime <- 12
eParsIn$collarNumYears <- 3

simBig <- getSimsNational()

scns <- expand.grid(
  obsYears = 10, projYears = 10, collarCount = 100, cowMult = 2, collarInterval = 2,
  assessmentYrs = 1, iAnthro = 0, obsAnthroSlope = 0, projAnthroSlope = 0, 
  sQuantile = c(0.1, 0.5, 0.9), rQuantile = c(0.1, 0.5, 0.9), N0 = 1000
)
scResults <- runScnSet(
  scns, eParsIn, simBig, getKSDists = FALSE,
  # only set to speed up vignette. Normally keep defaults.
  Niter = 150, Nburn = 100)

plotRes(scResults,
        c("Recruitment", "Adult female survival", "Population growth rate"), 
        facetVars = c("rQuantile", "sQuantile"))

```

### Modifying Bayesian model priors
By default `caribouBayesianPM()` calls `getPriors()` internally to set the priors for the Bayesian model based on the national model and default uncertainty modifiers that have been calibrated to fit the national model while allowing for deviations based on local data. The parameters to `getPriors()` affect the level of uncertainty around the coefficients from the national model. If uncertainty in the national model is higher then the projections can deviate more from the national model if the observed data does. See `getPriors()` for details.  

## Troubleshooting
The national model results are cached if the default values are used. This cache can be updated by running `getSimsNational(forceUpdate = TRUE)`

<!--
## Use the Bayesian demographic projection app

In addition to performing Bayesian demographic projections in R you can also use our shiny app to run similar analyses from a Graphical User Interface. To launch the app you will first need to install the package from GitHub.

```{r eval=FALSE}
#install.packages("remotes")
remotes::install_github("LandSciTech/BayesianCaribouDemographicProjection")
```

Then you can call `demographicProjectionApp()` to launch the app in your default browser. The app includes options to modify the disturbance scenario, the simulated observations and the model priors similar to those described above. Detailed instructions are included on the first page. -->


