---
title: "Bayesian Demographic Projection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Demographic Projection}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 3.5
)
```

```{r setup}
library(caribouMetrics)
library(ggplot2)
library(dplyr)

theme_set(theme_bw())
```

`caribouMetrics` includes several functions designed to allow users to explore the impacts of various factors that affect projections of caribou population dynamics using the Johnson et al. (2020) national models as a starting point.

A Bayesian model is fit that uses the national model as the priors and is updated based on a simulated or observed data set. The priors from the national model can also be modified to allow greater uncertainty or reflect different expectations.

### Using real observed data

```{r load-obs}
survObs <- read.csv(system.file("extdata/simSurvData.csv", package = "caribouMetrics"))
ageRatioObs <- read.csv(system.file("extdata/simAgeRatio.csv", package = "caribouMetrics"))
distObs <- read.csv(system.file("extdata/simDisturbance.csv", package = "caribouMetrics"))

survObs %>% group_by(Year) %>% 
  summarise(n_collars = n(),
            n_die = sum(event)) %>% 
  ggplot(aes(Year))+
  geom_line(aes(y = n_collars))+
  geom_col(aes(y = n_die))+
  labs(y = "Number of Collars (line) and deaths (bars)")

ageRatioObs %>% 
  ggplot(aes(Year, Count, fill = Class))+
  geom_col(position = "dodge")

distObs %>% ggplot(aes(Year, Anthro))+
  geom_line()

```

```{r mod-real}
mod_real <- caribouBayesianIPM(survData = survObs, ageRatio = ageRatioObs, 
                              disturbance = distObs, 
                              Niter = 150, Nburn = 100)

mod_tbl <- getOutputTables(mod_real, 
                           simNational = getSimsNational(),
                           getKSDists = FALSE)

plotRes(mod_tbl,
        c("Recruitment", "Adult female survival", "Population growth rate"),
         labFontSize = 10)

```

From these graphs we can see that this local population seems to have slightly higher demographic rates than would have been predicted by the national model alone. Note that the population's response to anthropogenic disturbance in completely determined by the national model since there was 0% disturbance during the observation period.  

### Simulate observed data
We will simulate observations from a typical boreal caribou monitoring program that includes radio collars on some cows and annual aerial surveys to count the number of calves with the tracked cows. To run the simulations we need a data frame of the number of collars deployed in each year and a data frame of cow counts in each year. In addition we will provide parameters to create a disturbance scenario and set other parameters...

```{r sim-obs}

# decreasing cow counts each year as anthropogenic disturbance increases. This
# only affects the absolute number of cows in ageRatio but not the ratio and not
# survival estimates. These data frames only affect the amount of information
# and so the variation between the observations and the "True" population.

# TODO: I think this would be more clear if we only supplied freqStartsByYear
# and cmult so that custom collaring programs could be accounted for but there
# wouldn't be any expectation that demographics are reflected here.

cow_tbl <- data.frame(Year = 2004:2023,
                      Class = "cow",
                      Count = 1000 - ceiling(seq(from = 0, to = 900, length.out = 20)))

collar_dep_tbl <- data.frame(Year = 2004:2023,
                             numStarts = 50)

scn_params <- getScenarioDefaults(obsAnthroSlope = 2, projAnthroSlope = 3,
                                  obsYears = 20, projYears = 20)

# TODO: change so cowCounts and freqStartsByYear are not required and can
# instead be set with getScenarioDefaults's cowCount and collarCount. 
sim_obs <- simulateObservations(
  scn_params, cowCounts = cow_tbl, freqStartsByYear = collar_dep_tbl,
  printPlot = TRUE
)


sim_obs$simSurvObs %>% group_by(Year) %>% 
  summarise(n_collars = n(),
            n_die = sum(event)) %>% 
  ggplot(aes(Year))+
  geom_line(aes(y = n_collars))+
  geom_col(aes(y = n_die))+
  labs(y = "Number of Collars (line) and deaths (bars)")

sim_obs$ageRatioOut %>% 
  ggplot(aes(Year, Count, fill = Class))+
  geom_col(position = "dodge")

sim_obs$simDisturbance %>% ggplot(aes(Year, Anthro))+
  geom_line()
```

```{r mod-sim-obs}
# TODO default start and end year should be set by disturbance data for IPM,
# output tables and tabAllRes
mod_sim <- caribouBayesianIPM(survData = sim_obs$simSurvObs,
                              ageRatio = sim_obs$ageRatioOut, 
                              disturbance = sim_obs$simDisturbance, 
                              startYear = min(sim_obs$simDisturbance$Year), 
                              endYear = max(sim_obs$simDisturbance$Year), 
                              Niter = 150, Nburn = 100)
simNational <- getSimsNational()
mod_sim_tbl <- getOutputTables(mod_sim, simObsList = sim_obs,
                           simNational = simNational,
                           startYear = min(sim_obs$simDisturbance$Year),
                           endYear = max(sim_obs$simDisturbance$Year),
                           getKSDists = FALSE)

plotRes(mod_sim_tbl,
        c("Recruitment", "Adult female survival", "Population growth rate"),
        labFontSize = 10)
```


### Comparing many scenarios

```{r many-scns, fig.height=5}
eParsIn <- list()
eParsIn$cowCounts <- data.frame(
  Year = 1981:2023,
  Count = 100,
  Class = "cow"
)
eParsIn$freqStartsByYear <- data.frame(
  Year = 1981:2023,
  numStarts = 100
)
eParsIn$collarOnTime <- 1
eParsIn$collarOffTime <- 12
eParsIn$collarNumYears <- 3

# adjust recruitment for delayed age of first reproduction
adjustR <- T 

simBig <- getSimsNational(adjustR = adjustR)

scns <- expand.grid(
  obsYears = 10, projYears = 10, collarCount = 100, cowMult = 2, collarInterval = 2,
  assessmentYrs = 1, iAnthro = 0,
  tA = 0, obsAnthroSlope = 0, projAnthroSlope = 0, sQuantile = c(0.1, 0.5, 0.9), 
  rQuantile = c(0.1, 0.5, 0.9), N0 = 1000
)
scResults <- caribouMetrics:::runScnSet(scns, eParsIn, simBig,
                                        getKSDists = FALSE, Niter = 150, Nburn = 100)

plotRes(scResults,
        c("Recruitment", "Adult female survival", "Population growth rate"), 
        facetVars = c("rQuantile", "sQuantile"))

```


### Troubleshooting
The national model results are cached if the default values are used. This cache can but updated by running `getSimsNational(forceUpdate = TRUE)`




