---
title: "Caribou Demographic Rates and Trajectories"
output: 
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Caribou Demographic Rates and Trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
pkgdown:
  as_is: true
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

if (!rlang::is_installed("patchwork")) stop("Please install.packages('patchwork') to run this vignette")
```

# Demographic rates and trajectories from the national model

## Overview
Here we describe a demographic model with density dependence and interannual variability following @johnson2020 with modifications noted in  @hughes2025 and [Dyson et al. (2022)](https://doi.org/10.1101/2022.06.01.494350). Demographic rates vary with disturbance as estimated by @johnson2020. A detailed description of the model is provided in [@hughes2025 Section 2.4].

`getNationalCoefficients()` selects the regression coefficient values and standard errors for the desired model version (see `popGrowthTableJohnsonECCC` for options) and then samples coefficients from these Gaussian distributions for each replicate population.  

Next `estimateNationalRates()` is used to apply the sampled coefficients to the disturbance covariates to calculate expected recruitment and survival according to the beta regression models estimated by @johnson2020. Each population is optionally assigned to quantiles of the Beta error distributions for survival and recruitment. Using quantiles means that the population will stay in these quantiles as disturbance changes over time, so there is persistent variation in recruitment and survival among example populations.

Finally, we can use the estimated demographic rates to project population dynamics using a simple model with two age classes. Interannual variation in survival and recruitment is modelled using truncated Beta distributions.

```{r setup, results='hide',message=F, warning=T}
library(caribouMetrics)
# use local version on local and installed on GH
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all()
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)

theme_set(theme_bw())
pthBase <- system.file("extdata", package = "caribouMetrics")

figWidth <- 8; figHeight <- 10

```

## Demographic outcomes in a simple case no variation in disturbance over time or among populations

A simple case for demographic projection is multiple stochastic projections from a single landscape that does not change over time. First we define a disturbance scenario with 40% anthropogenic disturbance and 2% fire disturbance. If we had spatial data for the disturbance in our area of interest we could use `disturbanceMetrics()` to directly calculate the disturbance. (See [Disturbance Metrics](https://landscitech.github.io/caribouMetrics/articles/Using_disturbanceMetrics.html) vignette for an example).
```{r make-dist}
disturbance <- data.frame(Anthro = 40, fire_excl_anthro = 2)
```

We begin by sampling coefficients for 500 replicate populations using default Johnson et al. (2020) models "M1" and "M4". The returned object is a list containing the coefficients and standard errors from the national model as well as the sampled coefficients and the quantiles that they have been assigned to.

```{r coefs}
popGrowthPars <- getNationalCoefficients(500)

head(popGrowthPars$coefSamples_Survival$coefSamples)
head(popGrowthPars$coefSamples_Survival$coefValues)
head(popGrowthPars$coefSamples_Survival$coefStdErrs)
head(popGrowthPars$coefSamples_Survival$quantiles)
```

Next we calculate sample demographic rates given sampled model coefficients and disturbance metrics for our example landscape, setting `returnSample = TRUE` so that the results returned contain a row for each sample in each scenario. We set the initial population size for each sample population to 100, and project population dynamics for 20 years using the `caribouPopGrowth` function with default parameter values. Anthropogenic disturbance is high on this example landscape, so the projected population growth rate for most sample populations is below 1, but there is variability in the model so few sample populations persist.

```{r simpleProjection, results='hide',message=F, warning=T}
rateSamples <- estimateNationalRates(
  covTable = disturbance,
  popGrowthPars = popGrowthPars,
  ignorePrecision = FALSE,
  returnSample = TRUE,
  useQuantiles = FALSE)     
rateSamples$N0 <- 100
demography <- cbind(rateSamples,
                    caribouPopGrowth(N = rateSamples$N0,
                                     numSteps = 20,
                                     R_bar = rateSamples$R_bar,
                                     S_bar = rateSamples$S_bar))

```

```{r plotSimpleDemography, echo=F, fig.width=figWidth,fig.height=5,fig.cap="Variation in demographic rates and outcomes among 500 sample populations after a 20 year projection. Anthropogenic disturbance is 40%, fire disturbance is 2%. 'lambda' is realized population growth rate in the final year, and 'lambdaE' is expected population growth rate without interannual variation, density dependence or demographic stochasticity. 'N' is the number of adult females, which includes new recruits 'n_recruits' and survivors 'surviving_adFemales' from the previous year. 'R_bar/S_bar' are the expected recruitment (calf:cow ratio) and survival rates, 'R_t/S_t' are the recruitment and survival rates in the final year, which are more variable because they include interannual variation. 'X_t' is the recruitment rate adjusted for sex ratio and (optionally) composition survey bias - see Hughes et al. (2025) for details."}
fds <- pivot_longer(demography, !(scnID:replicate) & !N0, names_to = "Metric",
                    values_to = "Amount")
d1 <- ggplot(fds, aes(x = as.factor(round(Anthro, 2)), y = Amount, 
                      colour = fire_excl_anthro)) +
  geom_violin(alpha = 0.4, color = "black") +
  geom_point(shape = 21, size = 2, position = position_jitterdodge()) +
  facet_wrap(~Metric, scales = "free") +
  theme(legend.position = "none") +
  xlab("Anthro")

plot(d1)
```

TO DO: Use consistent labels for outcomes throughout

TO DO: show and explain outcome if no N0 provided.

## Expected recruitment and survival: effects of disturbance and variation among populations

We can project demographic rates over a range of landscape conditions to recreate figures 3 and 5 from Johnson et al. [@johnson2020] and see the effects of changing disturbance on expected recruitment and survival. First we create a table of disturbance scenarios across a range of different levels of fire and anthropogenic disturbance.

```{r disturbanceCovariates, results='hide',message=F, warning=T}
covTableSim <- expand.grid(Anthro = seq(0, 90, by = 2), 
                           fire_excl_anthro = seq(0, 70, by = 10)) 
covTableSim$Total_dist = covTableSim$Anthro + covTableSim$fire_excl_anthro
```

We again sample coefficients from default models M1 and M4. The sample of 500 is used to calculate averages, while the sample of 35 is used to show variability among populations.

```{r popGrowthParameters, results='hide',message=F, warning=T}
# set seed so vignette looks the same each time
set.seed(34533)

popGrowthPars <- getNationalCoefficients(
  500,
  modelVersion = "Johnson",
  survivalModelNumber = "M1",
  recruitmentModelNumber = "M4",
  populationGrowthTable = popGrowthTableJohnsonECCC
)

popGrowthParsSmall <- getNationalCoefficients(
  35,
  modelVersion = "Johnson",
  survivalModelNumber = "M1",
  recruitmentModelNumber = "M4",
  populationGrowthTable = popGrowthTableJohnsonECCC
)
```

Next we calculate expected survival and recruitment rates given sampled model coefficients. For the smaller sample we set `returnSample = TRUE` so that the results returned contain a row for each sample in each scenario. Setting `useQuantiles = TRUE` assigns each sample population to a quantile of the regression model error distributions for survival and recruitment, allowing variation among populations to persist as disturbance changes. For the larger sample we set `returnSample = FALSE` to get summaries of mean and variation across the samples. 

Johnson et al's [-@johnson2020] Beta regression models estimate uncertainty about intercept and slope coefficients, and a precision parameter that describes variation among observations [@ferrari2004]. To show the importance of considering both these sources of variation we compare results with `ignorePrecision = TRUE` and `ignorePrecision = FALSE`. If the goal is to estimate and visualize nationally applicable demographic-disturbance relationships then the precision parameter may not be of interest (Figure \@ref(fig:parameterUncertaintyOnly)). If we are interested in the distribution of variation across the country it is essential to include both uncertainty about the regression coefficients and additional variation summarized by the precision parameter of the Beta regression model (Figure \@ref(fig:withPrecision)). 

```{r estimateNationalRates, results='hide',message=F, warning=T}
rateSamples <- estimateNationalRates(
  covTable = covTableSim,
  popGrowthPars = popGrowthParsSmall,
  ignorePrecision = FALSE,
  returnSample = TRUE,
  useQuantiles = TRUE
)

rateSummaries <- estimateNationalRates(
  covTable = covTableSim,
  popGrowthPars = popGrowthPars,
  ignorePrecision = FALSE,
  returnSample = FALSE,
  useQuantiles = FALSE
)

rateSummariesIgnorePrecision <- estimateNationalRates(
  covTable = covTableSim,
  popGrowthPars = popGrowthPars,
  ignorePrecision = TRUE,
  returnSample = FALSE,
  useQuantiles = FALSE
)
```

```{r parameterUncertaintyOnly, echo=F, fig.width=figWidth,fig.height=5,fig.cap="Variation in expected survival and recruitment with disturbance. Bands (2.5% and 97.5% quantiles of 500 samples) show variation due to uncertainty about intercept and slope coefficients in Johnson et al's [-@johnson2020] Beta regression models."}
rateSamples$rep = as.factor(rateSamples$replicate)
levels(rateSamples$rep) = sample(unique(rateSamples$replicate), replace = FALSE)
rateSamples$rep = as.character(rateSamples$rep)
rateSamples$fullGrp = paste(rateSamples$rep, rateSamples$fire_excl_anthro)

base1 <- ggplot(data = rateSummariesIgnorePrecision, 
                aes(x = Anthro, y = S_bar, ymin = S_PIlow, ymax = S_PIhigh))+
  geom_line(colour = "black", linewidth = 2)+
  geom_ribbon(alpha = 0.25)+
  xlab("Anthropogenic Disturbance (%)")+
  ylab("Adult Female Survival")+
  scale_y_continuous(limits = c(0.65, 1))+
  theme(legend.position = "none") +
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))

plot_recruitment3 <- ggplot(
  data = rateSummariesIgnorePrecision, 
  aes(x = Anthro, y = R_bar * 100, group = fire_excl_anthro, colour = fire_excl_anthro, type = fire_excl_anthro, 
      ymin = R_PIlow * 100, ymax = R_PIhigh * 100, fill = fire_excl_anthro)
  )+
  geom_line(linewidth = 2)+
  geom_ribbon(alpha = 0.15, colour = NA)+
  scale_fill_gradient(low = "green", high = "red")+
  scale_colour_gradient(low = "green", high = "red")+
  scale_y_continuous(limits = c(0, 60), breaks = c(0, 10, 20, 30, 40, 50, 60))+
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))+
  xlab("Anthropogenic disturbance (%)")+
  ylab("Recruitment (calves per 100 cows)")

base1 / plot_recruitment3
```
```{r withPrecision, echo=F, fig.width=figWidth,fig.height=5,fig.cap="Variation in expected survival and recruitment with disturbance. Bands (2.5% and 97.5% quantiles of 500 samples) include both uncertainty about the regression coefficients and additional variation summarized by the precision parameter of Johnson et al's [-@johnson2020] Beta regression models. Faint coloured lines show example trajectories of expected demographic rates in sample populations, assuming each sample population is randomly distributed among quantiles of the beta distribution, and each population remains in the same quantile of the Beta distribution as disturbance changes."}
base1 <- ggplot(data = rateSummaries, 
                aes(x = Anthro, y = S_bar, ymin = S_PIlow, ymax = S_PIhigh))+
  geom_line(data = subset(rateSamples, fire_excl_anthro == 0), linewidth = 0.5, 
            aes(x = Anthro, y = S_bar, group = rep, colour = rep), inherit.aes = FALSE)+
  geom_line(colour = "black", linewidth = 2)+
  geom_ribbon(alpha = 0.25)+
  xlab("Anthropogenic Disturbance (%)")+
  ylab("Adult Female Survival")+
  scale_y_continuous(limits = c(0.65, 1))+
  theme(legend.position = "none")+
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))

plot_recruitment3 <- ggplot(data = rateSummaries, 
                            aes(x = Anthro, y = R_bar * 100,  group = fire_excl_anthro,
                                colour = fire_excl_anthro, type = fire_excl_anthro))+
  geom_line(data = rateSamples,  
            aes(x = Anthro, y = R_bar * 100, group = fullGrp, colour = fire_excl_anthro),
            linewidth = 0.5, alpha = 0.2, inherit.aes = FALSE)+
  geom_line(linewidth = 2)+
  scale_colour_gradient(low = "green", high = "red")+
  scale_y_continuous(limits = c(0, 60),
                     breaks = c(0, 10, 20, 30, 40, 50, 60))+
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))+
  xlab("Anthropogenic disturbance (%)")+
  ylab("Recruitment (calves per 100 cows)")
base1 / plot_recruitment3
```

## Projection of population growth over time on a changing landscape: workflow details

In this example, we project 35 sample populations for 50 years on a landscape where the anthropogenic disturbance footprint is increasing by 5% per decade (Figure \@ref(fig:changeOverTime)). 

```{r projectionOverTime, results='hide',message=F, warning=T}
numTimesteps <- 5
stepLength <- 10
N0 <- 100
AnthroChange <- 5 #For illustration assume 5% increase in anthropogenic disturbance footprint each decade

# at each time,  sample demographic rates and project, save results
pars <- data.frame(N0 = N0)
for (t in 1:numTimesteps) {
  #t=1
  covariates <- disturbance
  covariates$Anthro <- covariates$Anthro + AnthroChange * (t - 1)

  rateSamples <- estimateNationalRates(
    covTable = covariates,
    popGrowthPars = popGrowthParsSmall,
    ignorePrecision = FALSE,
    returnSample = TRUE,
    useQuantiles = TRUE
  )
  if (is.element("N", names(pars))) {
    pars <- subset(pars, select = c(replicate, N))
    names(pars)[names(pars) == "N"] <- "N0"
  }
  pars <- merge(pars, rateSamples)
  pars <- cbind(pars, 
                caribouPopGrowth(pars$N0,
                                 R_bar = pars$R_bar, S_bar = pars$S_bar,
                                 numSteps = stepLength))

  # add results to output set
  fds <- subset(pars, select = c(replicate, Anthro, N, S_bar,S_t, R_bar,R_t,lambdaE,lambda))
  fds$replicate <- as.numeric(gsub("V", "", fds$replicate))
  #names(fds) <- c("Replicate", "anthro", "survival", "recruitment", "N", "lambda", "lambda_bar")
  fds <- pivot_longer(fds, !replicate, names_to = "MetricTypeID", values_to = "Amount")
  fds$Timestep <- t * stepLength
  if (t == 1) {
    popMetrics <- fds
  } else {
    popMetrics <- rbind(popMetrics, fds)
  }
}

popMetrics$MetricTypeID <- as.character(popMetrics$MetricTypeID)
popMetrics$Replicate <- paste0("x", popMetrics$replicate)
# popMetrics <- subset(popMetrics, !MetricTypeID == "N")
```

```{r changeOverTime, echo=F, fig.width=figWidth,fig.height=7,fig.cap="TO DO."}
base1 <- ggplot(data = popMetrics, aes(x = Timestep, y = Amount, colour = Replicate, 
                                       group = Replicate)) +
  geom_line() +
  facet_wrap(~MetricTypeID, scales = "free") +
  xlab("Time") +
  theme(legend.position = "none")
base1
```

## Using the trajectoriesFromNational wrapper function to project population growth

The examples above show steps in the demographic modeling workflow. We also provide a `trajectoriesFromNational` wrapper function that samples the coefficients from the National model, calculates demographic rates given those coefficients and the level of disturbance, projects population growth using `caribouPopGrowth`, and returns summaries of the demographic rates (ref workflow diagram). If the disturbance scenario includes a Year column `trajectoriesFromNational` projects population growth over time, and also returns sample demographic trajectories. If year is not provided, population growth is projected for one year, and sample trajectories are not returned.  

TO DO: add trajectoriesFromNational workflow diagram

```{r results='hide',message=F, warning=T}
natTraj <- trajectoriesFromNational(replicates = 500, 
                                    disturbance = covTableSim, interannualVar = FALSE, 
                                    useQuantiles = TRUE)

# add year to return samples
natTraj35 <- trajectoriesFromNational(replicates = 35, 
                                    disturbance = covTableSim %>% 
                                      mutate(Year = Anthro+100*fire_excl_anthro), 
                                    returnSamples = TRUE, interannualVar = FALSE, 
                                    useQuantiles = TRUE)
```

```{r withPrecisionWrapper, echo=F, fig.width=figWidth,fig.height=5,fig.cap="TO DO."}
base2 <- ggplot(data = natTraj$summary %>% filter(MetricTypeID == "survival", fire_excl_anthroID == 0) ,
                aes(x = AnthroID, y = Mean, ymin = lower, ymax = upper))+
   geom_line(data = natTraj35$samples%>% 
               filter(MetricTypeID == "survival", fire_excl_anthroID == 0),
             linewidth = 0.5,
            aes(x = AnthroID, y = Amount, group = Replicate, colour = Replicate),
            inherit.aes = FALSE)+
  geom_line(colour = "black", linewidth = 2)+
  geom_ribbon(alpha = 0.25)+
  xlab("Anthropogenic Disturbance (%)")+
  ylab("Adult Female Survival")+
  scale_y_continuous(limits = c(0.65, 1))+
  theme(legend.position = "none")+
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))
plot_recruitment4 <- ggplot(data = natTraj$summary %>% filter(MetricTypeID == "recruitment"),
                            aes(x = AnthroID, y = Mean * 100,  group = fire_excl_anthroID,
                                colour = fire_excl_anthroID, type = fire_excl_anthroID))+
  geom_line(data = natTraj35$samples%>% 
               filter(MetricTypeID == "recruitment"),
            aes(x = AnthroID, y = Amount*100, group =interaction(Replicate, fire_excl_anthroID),
                colour = fire_excl_anthroID),
            linewidth = 0.5, alpha = 0.2, inherit.aes = FALSE)+
  geom_line(linewidth = 2)+
  scale_colour_gradient(low = "green", high = "red")+
  scale_y_continuous(limits = c(0, 60),
                     breaks = c(0, 10, 20, 30, 40, 50, 60))+
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))+
  xlab("Anthropogenic disturbance (%)")+
  ylab("Recruitment (calves per 100 cows)")
plot_lambda <- ggplot(data = natTraj$summary %>% filter(MetricTypeID == "recruitment"),
                            aes(x = AnthroID, y = Mean * 100,  group = fire_excl_anthroID,
                                colour = fire_excl_anthroID, type = fire_excl_anthroID))+
  geom_line(data = natTraj35$samples%>% 
               filter(MetricTypeID == "recruitment"),
            aes(x = AnthroID, y = Amount*100, group =interaction(Replicate, fire_excl_anthroID),
                colour = fire_excl_anthroID),
            linewidth = 0.5, alpha = 0.2, inherit.aes = FALSE)+
  geom_line(linewidth = 2)+
  scale_colour_gradient(low = "green", high = "red")+
  scale_y_continuous(limits = c(0, 60),
                     breaks = c(0, 10, 20, 30, 40, 50, 60))+
  scale_x_continuous(limits = c(0, 90), breaks = c(0, 20, 40, 60, 80))+
  xlab("Anthropogenic disturbance (%)")+
  ylab("Recruitment (calves per 100 cows)")


base2 / plot_recruitment4
```

Using the same scenario with anthropogenic disturbance footprint increasing by 5% per decade, we 
can also produce projections over a changing landscape with `trajectoriesFromNational`. 

```{r results='hide',message=F, warning=T}
disturbance2 = data.frame(step = 0:4) %>% bind_cols(disturbance) %>% 
  mutate(Anthro = Anthro + AnthroChange * step, 
         Year = step * 10)

# set seed so vignette looks the same each time
set.seed(54545)

popMetrics2 <- trajectoriesFromNational(disturbance = disturbance2, replicates = 35, 
                                        useQuantiles = TRUE,N0 = 100, numSteps = 10)

popMetrics2 <- popMetrics2$samples %>% 
  filter(MetricTypeID %in% c("Anthro", "N", "Sbar","survival","Rbar","recruitment", "lambda_bar", "lambda"))
```

```{r changeOverTime2, echo=F, fig.width=figWidth,fig.height=7,fig.cap="TO DO."}
proj <- ggplot(data = popMetrics2, 
               aes(x = Timestep, y = Amount, colour = Replicate,
                   group = Replicate)) +
  geom_line() +
  facet_wrap(~MetricTypeID, scales = "free") +
  xlab("Time") +
  theme(legend.position = "none")
proj
```


## References
  Dyson, M., Endicott, S., Simpkins, C., Turner, J. W., Avery-Gomm, S.,
  Johnson, C. A., Leblond, M., Neilson, E. W., Rempel, R., Wiebe, P. A.,
  Baltzer, J. L., Stewart, F. E. C., & Hughes, J. (2022). Existing
  caribou habitat and demographic models need improvement for Ring of Fire
  impact assessment: A roadmap for improving the usefulness, transparency,
  and availability of models for conservation.
  <https://doi.org/10.1101/2022.06.01.494350>
  
  Novomestky F, Nadarajah S (2016) Package ‘truncdist.’ Version 1.0-2URL
  <https://CRAN.R-project.org/package=truncdist>

  
