---
title: "caribouHabitat"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{caribouHabitat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
This vignette gives details on the use of the `caribouHabitat` function. See `vignette("Getting_started", package = "caribouMetrics")` for an overview of the function. This function implements the caribou resource selection probability functions (RSPF) described in [Hornseth and Rempel, 2016](https://doi.org/10.1139/cjz-2015-0101) in R. They were previously written in LSL and the goal of this package is to make them more accessible to a wider range of users. 

```{r setup}
devtools::load_all()
pthBase <- system.file("extdata", package = "caribouMetrics")
```

## Simple Example
The example data set loaded below includes a small area in the Nipigon caribou range that we will use as an example. The PLC data are converted to resource types based on look up tables provided in the package 

```{r load data}
# load example data and classify plc into Resource Types
landCoverD = raster(file.path(pthBase, "landCover.tif")) %>% 
  reclassPLC()
eskerDras = raster(file.path(pthBase, "eskerTif.tif"))
eskerDshp = st_read(file.path(pthBase, "esker.shp"), quiet = TRUE)
natDistD = raster(file.path(pthBase, "natDist.tif"))
anthroDistD = raster(file.path(pthBase, "anthroDist.tif"))
linFeatDras = raster(file.path(pthBase, "linFeatTif.tif"))
projectPolyD = st_read(file.path(pthBase, "projectPoly.shp"), quiet = TRUE)
linFeatDshp = st_read(file.path(pthBase, "linFeat.shp"), quiet = TRUE)
roadsD = st_read(file.path(pthBase, "roads.shp"), quiet = TRUE)
railD = st_read(file.path(pthBase, "rail.shp"), quiet = TRUE)
utilitiesD = st_read(file.path(pthBase, "utilities.shp"), quiet = TRUE)
```

`caribouHabitat` will prepare the data, process it into the explanatory variables in the Hornseth and Rempel RSPF models and then calculate the probability of habitat use in each season at each location. The function can be run in several different ways, but the simplest is to provide spatial objects for each data set. 

```{r caribouHabitat1}
carHab1 <- caribouHabitat(
  landCover = landCoverD , 
  esker = eskerDras, 
  natDist = natDistD, 
  anthroDist = anthroDistD, 
  linFeat = linFeatDras, 
  projectPoly = projectPolyD,
  caribouRange = "Nipigon"
)
```

```{r plot1, fig.width=8, fig.height=5}
plot(carHab1, season = c("Fall", "Summer"))
```

## Example with padding

In most cases you will have a project area that is smaller than the environmental data sets available. For our example we will create this project area by selecting a rectangle inside our example data set. 

```{r create projPoly}
ext <- raster::extent(landCoverD) - 30000

projectPolyD2 <- st_bbox(ext) %>% st_as_sfc() %>% st_as_sf() %>% st_set_crs(st_crs(landCoverD))
```

The information that is outside our project area is still useful for preventing edge effects in our results. To use this data you can set `padProjPoly = TRUE` in the `caribouHabitat` call which will set a buffer around the project area based on the size of the moving window used to scale variables for that range. 

```{r caribouHabitat2}
carHab2 <- caribouHabitat(
  landCover = landCoverD, 
  esker = eskerDshp, 
  linFeat = linFeatDshp, 
  projectPoly = projectPolyD2,
  caribouRange = "Nipigon", 
  padProjPoly = TRUE
)
```

```{r inspect padded, fig.width=8, fig.height=5}
tmap::qtm(landCoverD, raster = "black")+
  tmapCatRast(carHab2@landCover, lbls = raster::levels(carHab2@landCover)[[1]][[2]], ttle = "PLC")+
  tmap::qtm(carHab2@projectPoly, fill = NULL, borders = "black")
```

The black area on the map above shows the extent of the original plc data, the raster shows the plc data that has been cropped to a buffer around projectPoly which is the black rectangle.  

```{r compare pad, fig.width=8, fig.height=5}
carHab3 <- caribouHabitat(
  landCover = landCoverD, 
  esker = eskerDshp, 
  linFeat = linFeatDshp, 
  projectPoly = projectPolyD2,
  caribouRange = "Nipigon"
)

carHab4 <- caribouHabitat(
  landCover = landCoverD, esker = eskerDshp, 
  linFeat = linFeatDshp,
  projectPoly = projectPolyD2,
  caribouRange = "Nipigon", 
  padProjPoly = FALSE, 
  padFocal = TRUE
)

st_area(projectPolyD2)


plot(carHab2, season = "Fall", title = "Pad with data outside project", 
     raster.breaks = c(0, 0.025, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 1), 
     layout.legend.outside = TRUE)+
  tmap::qtm(carHab3@projectPoly, fill = NULL, borders = "red", 
            shape.is.master = TRUE)

plot(carHab3, season = "Fall", title = "No padding", 
     raster.breaks = c(0, 0.025, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 1),
     layout.legend.outside = TRUE)+
  tmap::qtm(carHab3@projectPoly, fill = NULL, borders = "red", 
            shape.is.master = TRUE)

plot(carHab4, season = "Fall", title = "Pad with 0 outside project", 
     raster.breaks = c(0, 0.025, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 1),
     layout.legend.outside = TRUE)+
  tmap::qtm(carHab3@projectPoly, fill = NULL, borders = "red", 
            shape.is.master = TRUE)

```

The difference is large in this example because the project area is very small compared to the window size. The example with no padding is the default because it prevents users from relying on questionable results. The example padded with data outside the project polygon is the preferred option since it makes use of all available information but it is not the default because it should only be used when the data provided is larger than the project area. The version where cells are padded with 0s gives results but they can be misleading. Based on the land cover data in the previous map you can see that there is one an area with resource type LGOP near the border of projectPoly on the top right. LGOP has a strong positive effect on caribou habitat use so in the version padded with data cells on the top right edge have higher probability of use than in the version with 0s as padding. In general the predictions are lower around the edges in the version with 0s as padding because the cells outside the projectPoly are 0 when taking the weighted sum of values in the window. Therefore it is highly recommended to provide data for the area around the project and to use padProjPoly = TRUE.  

## Example with other inputs
As mentioned above `caribouHabitat` can be called in several different ways. You can provide the data as filepaths instead of spatial objects, you can provide the eskers and linear features as vector files which will be converted to rasters, and you can provide a list of sf objects to be combined into the linear features raster. 

```{r carbiouHabitat4}
carHab4 <- caribouHabitat(
  landCover = landCoverD, 
  esker = eskerDshp, 
  natDist = natDistD,
  anthroDist = anthroDistD,
  linFeat = list(roads = roadsD, rail = railD, utilities = utilitiesD),
  projectPoly = projectPolyD,
  caribouRange = "Nipigon"
)
```

## Example of updating an existing object
In some cases you may want to update an existing object with new data for one or more of the input data sets. This can be done using `updateCaribou`.

For example we can create several data sets that show roads expanding across our study are and what effect this will have on caribou habitat. Since the only dataset that is changing is the linear features it will be faster to use `updateCaribou` rather than `caribouHabitat`.

```{r updateCaribou, fig.width=6, fig.height=6}
# Create series of data sets
ext <- raster::extent(linFeatDras)

height <- dim(linFeatDras)[1]/4

linFeatDras1 <- linFeatDras[1:height, , drop = FALSE] %>%
  raster::extend(linFeatDras, value = 0)
linFeatDras2 <- linFeatDras[1:(height*2), , drop = FALSE] %>%
  raster::extend(linFeatDras, value = 0)
linFeatDras3 <- linFeatDras[1:(height*3), , drop = FALSE] %>%
  raster::extend(linFeatDras, value = 0)

par(mfrow = c(2, 2))
plot(linFeatDras1)
plot(linFeatDras2)
plot(linFeatDras3)
plot(linFeatDras)

# Run caribouHabitat to process all the data once
# Use Pagwachuan range because roads have a larger effect in that model
carHabLF1 <- caribouHabitat(
  landCover = landCoverD,
  esker = eskerDras,
  natDist = natDistD, 
  anthroDist = anthroDistD, 
  linFeat = linFeatDras1, 
  projectPoly = projectPolyD, 
  caribouRange = "Missisa", 
  padFocal = TRUE
)

# Run updateCaribou with the new linFeat raster
carHabLF2 <- updateCaribou(carHabLF1, newData = list(linFeat = linFeatDras2))

# compare results
plot(carHabLF1, season = "Spring", title = "linFeat 1", raster.breaks = 0:10/10, 
     legend.outside = TRUE)
plot(carHabLF2, season = "Spring", title = "linFeat 2", raster.breaks = 0:10/10, 
     legend.outside = TRUE)


```

You can see that there is a lower probability of caribou habitat use when there are more roads (linFeat 2).

## Example using Iteration

To go through a large number of updated data sets we can use the `purrr` package to iterate over different sets of data

```{r, fig.width=8, fig.height=5}

linFeatScns <- lst(linFeatDras2, linFeatDras3, linFeatDras)

# using carHabLF1 created above we run the update for all three datasets

linFeatResults <- purrr::map(
  linFeatScns,
  ~updateCaribou(carHabLF1, newData = list(linFeat = .x))
)

purrr::walk2(c(carHabLF1, linFeatResults), 
             c("linFeatDras1", names(linFeatResults)), 
             ~plot(.x, season = "Spring", title = .y, raster.breaks = 0:10/10,
                   legend.outside = TRUE) %>% print())


```

It is also possible to change just the `caribouRange` attribute without recomputing the entire function using the `updateCaribou` function. In the example below the coefRange is changed from Nipigon to Missisa and the predictions are re-calculated using the coefficients from the Missisa model

```{r update, fig.width=8, fig.height=5}
carHab1@attributes$caribouRange$coefRange <- "Missisa"

misCoef <- updateCaribou(carHab1)

plot(misCoef)
```

## Example with different coefficients
Different coefficients can be supplied using the `coefTable` argument and if those coefficients are standardized the data can be standarized by supplying the argument `doScale = TRUE`.
```{r}
carHabStd <- caribouHabitat(
  landCover = landCoverD,
  esker = eskerDras,
  natDist = natDistD, 
  anthroDist = anthroDistD, 
  linFeat = linFeatDras, 
  projectPoly = projectPolyD, 
  caribouRange = "James Bay", 
  coefTable = coefTableStd,
  doScale = TRUE
)
```

## Binary usage

You can also determine a binary high or low caribou use result either by season or for the range as a whole using `calcBinaryUse`

```{r binUse, fig.width=8, fig.height=5}
binCarHab1 <- calcBinaryUse(carHab1)
binCarHab1Seasons <- calcBinaryUse(carHab1, bySeason = TRUE)
tmap::qtm(binCarHab1)
tmap::qtm(binCarHab1Seasons)
```


