---
title: "Simulation of monitoring scenarios informed by local data"
output: 
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation of monitoring scenarios informed by local data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
pkgdown:
  as_is: true
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

if (!rlang::is_installed("patchwork")) stop("Please install.packages('patchwork') to run this vignette")
```

## Examples of initial Bayesian models informed by local data and national-demographic disturbance relationships

We simulate monitoring scenarios informed by example data from the `bboudata::bbousurv_a` and `bboudata::bbourecruit_a` data sets:

* Informative local data: Observations from 2010 to 2016.
* Limited local data: Observations from 2014 to 2016.

To examine and compare predictions for unobserved years (2018 to 2021) we add missing data.

Our starting points for analysis of monitoring scenarios are Bayesian Beta models informed by local data and prior knowledge of national demographic-disturbance relationships. See compare vignette for details. We examine the influence of prior information from national-demographic disturbance relationships in low disturbance (5% anthropogenic, 5% fire) and high disturbance (80% anthropogenic, 10% fire) scenarios.

```{r setup}
library(caribouMetrics)
# use local version on local and installed on GH
if(requireNamespace("devtools", quietly = TRUE)) devtools::load_all()
library(bboudata)
library(dplyr)
library(ggplot2)

figWidth <- 8
figHeight <- 10

#Note - set niters to 100 to run quickly when testing. Set to 1000 for complete results.
niters <- 100

useSaved <- T #option to skip slow step of fitting bboutools model
surv_data = bboudata::bbousurv_a %>% filter(Year > 2010)
surv_data_add = expand.grid(Year=seq(2017,2025),Month=seq(1:12),PopulationName=unique(surv_data$PopulationName))
surv_data=merge(surv_data,surv_data_add,all.x=T,all.y=T)
surv_data$StartTotal[is.na(surv_data$StartTotal)]=1

recruit_data=bboudata::bbourecruit_a %>% filter(Year > 2010)
recruit_data_add = expand.grid(Year=seq(2017,2025),PopulationName=unique(recruit_data$PopulationName))
recruit_data=merge(recruit_data,recruit_data_add,all.x=T,all.y=T)
recruit_data$Month[is.na(recruit_data$Month)]=3;recruit_data$Day[is.na(recruit_data$Day)]=15

surv_dataNone <- surv_data %>% filter(Year > 2017)
recruit_dataNone <- recruit_data %>% filter(Year > 2017)

surv_dataLimited <- surv_data %>% filter(Year > 2014)
recruit_dataLimited <- recruit_data %>% filter(Year > 2014)

disturbance = data.frame(Year = unique(surv_data$Year), Anthro = 5, fire_excl_anthro = 5)
betaLimited <- bbouMakeSummaryTable(surv_dataLimited, recruit_dataLimited, disturbance=disturbance,niters=niters)
betaInformative <- bbouMakeSummaryTable(surv_data, recruit_data, disturbance=disturbance,niters=niters)
simInformative <- getSimsInitial(betaInformative)
simLimited <- getSimsInitial(betaLimited)

disturbanceHigh <- data.frame(Year = disturbance$Year,Anthro = 80,fire_excl_anthro = 10)
betaLimitedHigh <- bbouMakeSummaryTable(surv_dataLimited, recruit_dataLimited, disturbance=disturbanceHigh)
simLimitedHigh <- getSimsInitial(betaLimitedHigh)
```

## Simulating additional monitoring of plausible example trajectories from fitted Bayesian models

We begin by specifying monitoring scenario parameters, and selecting an example trajectory with a relatively low expected growth rate (i.e. in the 1st percentile of the distribution of mcmc samples from the initial model). In this example, we specify 5 additional years of monitoring, with a target of 30 collars per year, and 6 cows per collared cow in the composition surveys. We then simulate additional years of monitoring, combine the simulated and observed data, and reanalyze the combined data. Methods for simulating monitoring of example trajectories are described in @hughes2025.

```{r Example monitoring scenario, results='hide'}
scns=list()
scns$lQuantile=0.01 #select example trajectories with low growth rate
correlateRates = T #Force correlation among demographic rates to examine extreme cases
scns$N0 = NA
scns$projYears = 3
scns$curYear = max(simInformative$recruit_data$Year)- scns$projYears
scns$obsYears = scns$curYear - min(simInformative$recruit_data$Year)
scns$cowMult = 6
scns$collarCount = 30

trajectories <- subset(simInformative$samples,LambdaPercentile == round(scns$lQuantile*100))
trajectories <- subset(trajectories,Replicate==sample(unique(trajectories$Replicate),1))

oo <- simulateObservations(getScenarioDefaults(scns), trajectories,
                           surv_data = simInformative$surv_data, recruit_data=simInformative$recruit_data)
informativeMoreMonitoring <- caribouBayesianPM(surv_data = oo$simSurvObs, recruit_data = oo$simRecruitObs, disturbance=disturbance)
out_tbls <- getOutputTables(informativeMoreMonitoring, simInitial = simInformative)
typeLabsI <- c("More", "Informative")

#Use runScnSet to select trajectory, simulate observations, fit new model and summarize results for the case when initial data is limited.
limitedMoreMonitoring = runScnSet(scns,simLimited) 
typeLabsL <- c("More", "Limited")

limitedMoreHigh = runScnSet(scns,simLimitedHigh) 
#TO DO: fix discrepancies between these two methods.
#TO DO: change dot labels. "True simulated" should be "trajectory", and "observed" should be "observation".
```

```{r fig-plot1, fig.width=figWidth,fig.height=10,fig.cap="Observed (2010-2017) and simulated (2018-2022) survival data for an example trajectory with a relatively low expected growth rate (i.e. in the 1st percentile of the distribution of mcmc samples from the initial model)."}
plotSurvivalSeries(oo$simSurvObs)
#TO DO: fix data gap
```

```{r fig-plot2, fig.width=figWidth,fig.height=10,fig.cap="95% prior predictive intervals from Beta models with and without additional monitoring (5 years) of an example trajectory with informative initial data (5 years), low disturbance, and a relatively low expected growth rate (Figure 1)."}
rec <- plotRes(out_tbls, "Recruitment", typeLabels = typeLabsI)
surv <- plotRes(out_tbls, "Adult female survival", typeLabels = typeLabsI)
lam <- plotRes(out_tbls, "Population growth rate", typeLabels = typeLabsI,
               lowBound = 0.5, highBound = 1.5)
rec / surv / lam
```

```{r fig-plot3, fig.width=figWidth,fig.height=10,fig.cap="95% prior predictive intervals from Beta models with and without additional monitoring (5 years) of an example trajectory with limited initial data (2 years), low disturbance, and a relatively low expected growth rate (Figure 1)."}
rec <- plotRes(limitedMoreMonitoring, "Recruitment", typeLabels = typeLabsL)
surv <- plotRes(limitedMoreMonitoring, "Adult female survival", typeLabels = typeLabsL)
lam <- plotRes(limitedMoreMonitoring, "Population growth rate", typeLabels = typeLabsL,
               lowBound = 0.5, highBound = 1.5)
rec / surv / lam
```

```{r fig-plot4, fig.width=figWidth,fig.height=10,fig.cap="95% prior predictive intervals from Beta models with and without additional monitoring (5 years) of an example trajectory with limited initial data (2 years), high disturbance, and a relatively low expected growth rate (Figure 1)."}
rec <- plotRes(limitedMoreHigh, "Recruitment", typeLabels = typeLabsL)
surv <- plotRes(limitedMoreHigh, "Adult female survival", typeLabels = typeLabsL)
lam <- plotRes(limitedMoreHigh, "Population growth rate", typeLabels = typeLabsL,
               lowBound = 0.5, highBound = 1.5)
rec / surv / lam
```

In these examples, an additional 5 years of monitoring of a plausible low growth rate trajectory with 5 intial years of data does not provide additional clarity about whether the population is likely to decline or not in future (Figure \@ref(fig:fig-plot2)). In contrast, an additional 5 years of monitoring does substantially reduce uncertainty when there is less initial data and disturbance is low (Figure \@ref(fig:fig-plot3)). When disturbance is high, additional monitoring does less to reduce uncertainty because we have clearer prior expectations from the national model even when local data is limited (Figure \@ref(fig:fig-plot4)).







## References


