---
title: "Simulation of monitoring scenarios informed by local data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation of monitoring scenarios informed by local data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(caribouMetrics)
```


## Example data

We simulate monitoring scenarios informed by example data from the `bboudata::bbousurv_a` and `bboudata::bbourecruit_a` data sets:

* Informative local data: Observations from 2010 to 2016.
* Limited local data: Observations from 2014 to 2016.

To examine and compare predictions for unobserved years (2018 to 2021) we add missing data.

```{r setup}
library(caribouMetrics)
# use local version on local and installed on GH
if(requireNamespace("devtools", quietly = TRUE)) devtools::load_all()
library(bboudata)
library(bboutools)
library(dplyr)
library(ggplot2)

figWidth = 8
figHeight = 4

useSaved <- T #option to skip slow step of fitting bboutools model
bbouInformativeFile <- here::here("results/vignetteBbbouExample.rds")
bbouLimitedFile <- here::here("results/vignetteBbbouExample2.rds")

surv_data = bboudata::bbousurv_a %>% filter(Year > 2010)
surv_data_add = expand.grid(Year=seq(2017,2022),Month=seq(1:12),PopulationName=unique(surv_data$PopulationName))
surv_data=merge(surv_data,surv_data_add,all.x=T,all.y=T)
surv_data$StartTotal[is.na(surv_data$StartTotal)]=1

recruit_data=bboudata::bbourecruit_a %>% filter(Year > 2010)
recruit_data_add = expand.grid(Year=seq(2017,2022),PopulationName=unique(recruit_data$PopulationName))
recruit_data=merge(recruit_data,recruit_data_add,all.x=T,all.y=T)
recruit_data$Month[is.na(recruit_data$Month)]=3;recruit_data$Day[is.na(recruit_data$Day)]=15

surv_dataNone <- surv_data %>% filter(Year > 2017)
recruit_dataNone <- recruit_data %>% filter(Year > 2017)

surv_dataLimited <- surv_data %>% filter(Year > 2014)
recruit_dataLimited <- recruit_data %>% filter(Year > 2014)
```

## Analysis of simulated monitoring using the bbou model with limited local data and no disturbance information

The starting point for analysis is a Bayesian model with a random effect of year and no annual trend implemented in the bboutools R package [Dalgarno et al. 2025]( https://doi.org/10.21105/joss.07997). For brevity, we refer to that as the "bbou" model.

```{r compare beta and bbou with limited data.}
if(useSaved&file.exists(bbouLimitedFile)){
  bbouLimited <- readRDS(bbouLimitedFile)
}else{
  bbouLimited <- bbouMakeSummaryTable(surv_dataLimited,recruit_dataLimited,return_mcmc=T)
  if(dir.exists(dirname(bbouLimitedFile))){saveRDS(bbouLimited,bbouLimitedFile)}
}
simBbouLimited <-getSimsInitial(bbouLimited) 
```







## Simulation of monitoring using Beta model with no disturbance


## Simulation of monitoring using Beta model, low disturance scenario


## Simulation of monitoring using Beta model, high disturance scenario

## References
Dalgarno S, Boulanger J, Pearson A, et al (2025) bbousuite: A set of R packages to facilitate analysis of boreal caribou survival and recruitment data. Journal of Open Source Software 10:7997. https://doi.org/10.21105/joss.07997

Hughes J, Endicott S, Calvert AM, Johnson CA (2025) Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions. Ecological Informatics 87:103095. https://doi.org/10.1016/j.ecoinf.2025.103095


