---
title: "Disturbance Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disturbance Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(caribouMetrics)
library(dplyr)
library(terra)
library(sf)
library(ggplot2)
library(tidyr)
theme_set(theme_bw())
pthBase <- system.file("extdata", package = "caribouMetrics")
```

# `disturbanceMetrics()`
The `disturbanceMetrics()` function is used to calculate the metrics described in Table 52 of Environment Canada Scientific Assessment to Inform the Identification of Critical Habitat for Woodland Caribou (*Rangifer tarandus caribou*), Boreal Population, in Canada 2011 Update. The metrics included are:
* Fire: % non-overlapping fire 
* Anthro: % non-overlapping anthropogenic disturbance
* Total_dist: % total non-overlapping fire and anthropogenic disturbance
* fire_excl_anthro: % fire not overlapping with anthropogenic disturbance

`disturbanceMetrics()` uses several spatial data layers to calculate the percentage disturbance in an area: 

| Name (Argument) | Description|
|-------------------------------------------------------|------------------------------------------------------------------|
| Land cover (landCover) | A raster where 0 and NA values are assumed to be water and are omitted from the total area, defines the raster grid |
| Linear features (linFeat) | a raster, sf object, or list of these identifying the location of linear features (e.g. roads, rail) |
| Natural disturbance (natDist) | Cumulative natural disturbance (mostly fire) over the past 40 years |  
| Anthropogenic disturbance (anthroDist)| Cumulative anthropogenic disturbance over the past 40 years |
|Project polygon (projPoly)| An sf object containing polygon(s) of the study area(s)| 

The example data set loaded below includes a small area in the Nipigon caribou range that we will use as an example. Disturbance data sets can be converted from polygons of year of disturbance or time since disturbance using (`reclassDist`). In the example below fire data with polygons containing year of disturbance are converted to a presence absence raster of cumulative disturbance over the past 40 years.

```{r load data, results='hide',message=F, warning=T}
# load example data and classify cumulative natural disturbance
landCoverD = rast(file.path(pthBase, "landCover.tif")) 
natDistD <- sf::st_read(file.path(pthBase, "fireAFFES2020.shp")) %>% 
  reclassDist(endYr = 2020, numCumYrs = 40, template = landCoverD, 
              dateField = "FIRE_YEAR")


anthroDistD = rast(file.path(pthBase, "anthroDist.tif"))
linFeatDras = rast(file.path(pthBase, "linFeatTif.tif"))
projectPolyD = st_read(file.path(pthBase, "projectPoly.shp"), quiet = TRUE)
linFeatDshp = st_read(file.path(pthBase, "roads.shp"), quiet = TRUE)
roadsD = st_read(file.path(pthBase, "roads.shp"), quiet = TRUE)
railD = st_read(file.path(pthBase, "rail.shp"), quiet = TRUE)
utilitiesD = st_read(file.path(pthBase, "utilities.shp"), quiet = TRUE)
```

`disturbanceMetrics` will prepare the data and then calculate disturbance metrics used as predictor variables by Johnson et al. (2020). The function can be run in several different ways, and the simplest is to provide spatial objects for each input. 

```{r disturbanceMetricsRaster, results='hide',message=F, warning=T}
disturbance <- disturbanceMetrics(
  landCover=!is.na(landCoverD),
  natDist = natDistD,
  anthroDist = anthroDistD,
  linFeat = linFeatDras,
  projectPoly = projectPolyD
)
```

The `disturbanceMetrics` function returns an S4 object with the class `DisturbanceMetrics`. To access a data.frame with one column per metric use the `results` function.

```{r results3}
str(disturbance, max.level = 2, give.attr = FALSE)
results(disturbance)
```

```{r plotDisturbance, fig.width=8, fig.height=5}
plot(disturbance@processedData)
```

Multiple linear feature inputs can also be provided in vector form. See `vignette("caribouHabitat", package = "caribouMetrics")` for other data processing and input options. 

```{r disturbanceMetricsVector, results='hide',message=F, warning=T}
disturbanceV <- disturbanceMetrics(
  landCover=!is.na(landCoverD),
  natDist = natDistD,
  anthroDist = anthroDistD,
  linFeat = list(roads = roadsD, rail = railD, utilities = utilitiesD),
  projectPoly = projectPolyD
)
```

```{r plotDisturbanceVector, fig.width=8, fig.height=5}
plot(disturbanceV@processedData)
```