---
author: "Sarah Endicott"
output: html_document
---
<style>
.vscroll-plot {
    width: 1000px;
    height: 500px;
    overflow-y: scroll;
    overflow-x: hidden;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# `r rmdTitle`

`r rmdDescrip`

Plots on the left are from Rob Rempel's results and ours are on the right. I have summarised the raster output in to hexagons for comparison.

## Overall results
Compares the results from raw data to prediction so differences include the methods for converting inputs to explanatory variables.

### Compare Category 2 Habitat

```{r combinedCat2}
mapCombinedCat2
```

#### Proportion Category 2 
```{r}
tblCombinedCat2 %>% knitr::kable(digits = 3, format = "html", table.attr = "style='width:30%;'")
```



### Compare results maps

<div class="vscroll-plot">
```{r resmap, comment=FALSE, message=FALSE, warning=FALSE}
purrr::walk(difMapRes, print)
```
</div>

### Compare variables maps

<div class="vscroll-plot">
```{r varsmap, message=FALSE, warning=FALSE}
purrr::walk(difMapVars, print)
```
</div>

### Compare results graphs

```{r resgr, message=FALSE, warning=FALSE}
lmRvsEPlot
```

### Compare variables graphs

<div class="vscroll-plot">
```{r varsgr, message=FALSE, warning=FALSE}
purrr::walk(varCompGraphs, print)
```
</div>

## Final calculation
Compares the results when the explanatory varibales given by Rempel are used to calculate the probability of use to Rempel's results. Therefore this only really tests the calcRSP function. 

### Map of difference
Calculated as Rempel_result-Endicott_result

```{r difMapCalc, message=FALSE}
difMapCalc
```

### Distribution of differences

```{r difDistSeasonGraph}
difDistSeasonGraph
```

### Difference in Fall vs each variable 

```{r}
difFAVarsGraph
```

## Conclusion

`r rmdConclu`